---
title: "Interactive Presentation Ninja"
subtitle: "⚔<br/>with interactingan"
author: "Juan Cruz Rodriguez"
institute: "FAMAF - UNC"
date: "2020/01/10 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

# Connect to the app!

```{r echo=FALSE, message=FALSE, fig.align='center'}
library("interactingan")
# Will configure an empty app to the shinyapps server;
# and show a qr code that points to the interactingan app url
set_app()
```

---

# Which R package you use most?

`r interactingan::wordcloud("Most used R package", max_words = 1)`
```{r shiny-ggally}
library(shiny)

ui <- fluidPage(
    
    # Application title
    titlePanel("Looking at Music Trends (make a selection)"),
    
    # Sidebar layout with a input and output definitions
    sidebarLayout(
        # Inputs: Select variables to plot
        sidebarPanel(
            
            # Select variable for location
            selectInput(inputId = "x", 
                        label = "X Axis: ",
                        choices = c("Valence" = "valence", "Acousticness" = "acousticness",
                                    "Danceability" = "danceability", "Duration" = "duration_ms", 
                                    "Energy" = "energy", "Instrumentalness" = "instrumentalness",
                                    "Liveness" = "liveness", "Loudness" = "loudness", 
                                    "Popularity" = "popularity", "Speechiness" = "speechiness"),
                        selected = "acousticness",
                        multiple = FALSE),
            
            selectInput(inputId = "y",
                        label = "Y Axis: ",
                        choices = c("Valence" = "valence", "Acousticness" = "acousticness",
                                    "Danceability" = "danceability", "Duration" = "duration_ms", 
                                    "Energy" = "energy", "Instrumentalness" = "instrumentalness",
                                    "Liveness" = "liveness", "Loudness" = "loudness", 
                                    "Popularity" = "popularity", "Speechiness" = "speechiness"),
                        selected = "energy",
                        multiple = FALSE),
            
            checkboxInput(inputId = "mod",
                          label = "Line",
                          value = TRUE)
        ),
        
        #Output
        
        mainPanel(
            
            # Show plot
            plotOutput(outputId = "plot"),
            br(),        # a little bit of visual separation
            
        )
    )
)

# Define server function --------------------------------------------
server <- function(input, output) {
    
    output$plot <- renderPlot({
   p3 <- 
     spotify_by_year %>%
     mutate(
       constant = 0.02
     ) %>%
     ggplot(aes_string(x = input$x, y = input$y)) +
     geom_point(aes(text = paste("Year:",year))) +
     theme_proj +
          #need to add gridlines because tufte removes them
          theme(panel.grid.major = element_line(color = "gray90"),
                panel.grid.minor = element_line(color = "gray90"),
                legend.position = "none") +
          labs(x = input$x, y = input$y,
               caption = "Source: Spotify") +
       if(input$mod) {geom_smooth(method = "lm", se = FALSE)}
   
   
    p3
    })
}

# Create the Shiny app object ---------------------------------------
shinyApp(ui, server, options = list(height = 500))
```

---

# Have you ever used interactive slides?

`r poll("Do you use interactive slides?", c("Yes", "No", "Won't answer!"))`

---

# How much do you like the `interactingan` package?

`r rating("Rate interactingan:")`

---

# What do you think about the `interactingan` package?

`r question("Opinion about interactingan:")`

---

# Questions?

<!-- Show the audience question's pane -->
`r audience_questions()`

```{r echo=FALSE}
# Deploys the final interactingan app to the shinyapps server
deploy_interactions()
```
