---
title: "Music, Crime, and Happiness"
author: |
  | Tolga Atabas & Sherry Huang
output:
  xaringan::moon_reader:
    css: stat41-lab-theme.css
    lib_dir: libs
    nature:
      highlightStyle: github
      countIncrementalSlides: false
---

# Xaringan presentation

(Need to install the `xaringan` package)

+ [About](https://bookdown.org/yihui/rmarkdown/xaringan.html)
+ [Themes here](https://github.com/yihui/xaringan/wiki/Themes)
+ Want to get fancy with a custom theme? Try [xaringan themer](https://pkg.garrickadenbuie.com/xaringanthemer/articles/xaringanthemer.html)
+ Style edits: google "how to ______ in xaringan"

**Difference:** slides are broken up with `---`

---
# Introduction

<iframe width="560" height="315" src="https://www.youtube.com/embed/rnUSNbqtVJI" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

Music has strong effects on the brain, specifically on: memory, movement, and emotions. Is it possible to see trends between music and happiness, and music and crime?

---
# Music Dataset

| Variable         | Type | Description                                                         | Example    |
|------------------|------|---------------------------------------------------------------------|------------|
| mode             | int  | whether progression of track is major (1) or minor (0)              | 1          |
| year             | int  | the year                                                            | 2020       |
| acousticness     | num  | relative metric of track being acoustic                             | 0.2199309  |
| danceability     | num  | relative measurement of track being danceable                       | 0.6929043  |
| duration_ms      | num  | length of track in milliseconds                                     | 193728.4   |
| energy           | num  | energy of track                                                     | 0.6312316  |
| instrumentalness | num  | relative ratio of track being instrumental                          | 0.01637552 |
| liveness         | num  | relative duration of track sounding as a live performance           | 0.1785354  |
| loudness         | num  | relative loudness of track in the typical range [-60, 0] in decibel | -6.595067  |
| speechiness      | num  | relative length of track containing any kind of human voice         | 0.14138369 |
| tempo            | num  | overall estimated tempo of a track in beats per minute              | 124.2831   |
| valence          | num  | describes the musical positiveness conveyed by a track              | 0.5010478  |
| popularity       | num  | popularity                                                          | 64.3019704 |
| key              | int  | key track is in                                                     | 1          |

---
# World Happiness Dataset

| Variable                      | Type | Description                                                                                            | Example        |
|-------------------------------|------|--------------------------------------------------------------------------------------------------------|----------------|
| Country                       | chr  | country                                                                                                | Switzerland    |
| Region                        | chr  | region country is located in                                                                           | Western Europe |
| Happiness.Rank                | int  | ranking by happiness score                                                                             | 1              |
| Happiness.Score               | num  | happiness score [0, 10]                                                                                | 7.587          |
| Standard.Error                | num  | standard error of happiness score                                                                      | 0.03411        |
| Economy..GDP.per.Capita.      | num  | is in terms of Purchasing Power Parity                                                                 | 1.39651        |
| Family                        | num  | family and social support                                                                              | 1.34951        |
| Health..Life.Expectancy.      | num  | healthy life expectancy at birth                                                                       | 0.94143        |
| Freedom                       | num  | freedom to make life choices                                                                           | 0.66557        |
| Trust..Government.Corruption. | num  | perceptions of corruption                                                                              | 0.41978        |
| Generosity                    | num  | generosity of people in the country                                                                    | 0.29678        |
| Dystopia.Residual             | num  | relative to a hypothetical country called “Dystopia” (with lowest national averages of previous years) | 2.51738        |
| year                          | int  | year                                                                                                   | 2015           |
| mean                          | num  | average happiness score                                                                                | 5.375734       |

---
# Crime Dataset

([tablesgenerator.com](https://www.tablesgenerator.com) is great for making markdown tables)

---
# Exploring Spotify Dataset

```{r scatter, echo=FALSE, message = FALSE, warning = FALSE, dpi = 200}
library(ggplot2)
library(ggthemes)
library(tidyverse)
library(dplyr)
library(maps)
library(plotly)
library(readr)
library(ggridges)
theme_proj = theme_tufte() + 
  theme(plot.title.position = "plot")
spotify_by_year <- read_csv("data/data_by_year.csv")
ggplot(data = spotify_by_year) +
  geom_point(mapping = aes(x = year, y = valence)) +
  labs(title = "Happiness of Music Over Time",
       x = "Year",
       y = "Valence (0 to 1)",
       caption = "SOURCE: Spotify") +
  theme_proj +
  #need to add gridlines because tufte removes them
  theme(panel.grid.major = element_line(color = "gray90"),
        panel.grid.minor = element_line(color = "gray90"))
```

```{r eda-2, echo = FALSE, message = FALSE, warning = FALSE, dpi = 200}
data <- read.csv("data/data.csv")

data %>%
  mutate(explicit = ifelse(explicit == 0, "Not Explicit", "Explicit")) %>%
  ggplot(aes(x = danceability, y = explicit)) +
  geom_density_ridges(aes(fill = explicit, group = explicit), alpha = 0.7) +
  labs(title = "Danceability of a song and whether it is explicit or not",
       subtitle = "(danceability is a measure from 0 to 1)",
       y = element_blank(),
       x = "Danceability") +
  scale_fill_manual(values = c( "#ffbebd","#1a405F")) +
  scale_y_discrete(limits = c("Not Explicit", "Explicit")) +
  theme_proj +
  guides(fill = FALSE)
```

---
# Exploring World Happiness Dataset

```{r, echo = FALSE, message = FALSE, warning = FALSE}
#loading dataset of each country's ISO code (because plotly needs it to map countries)
# source: https://www.kaggle.com/andradaolteanu/iso-country-codes-global
country_code <- read_csv("data/wikipedia-iso-country-codes.csv")
happiness_2015 <- read.csv("data/2015.csv") %>% mutate(year = 2015)
happiness_2016 <- read.csv("data/2016.csv") %>% mutate(year = 2016)
happiness_2017 <- read.csv("data/2017.csv") %>% mutate(year = 2017)
happiness_2018 <- read.csv("data/2018.csv") %>% mutate(year = 2018)
happiness_2019 <- read.csv("data/2019.csv") %>% mutate(year = 2019)
happiness_2020 <- read.csv("data/2020.csv") %>% mutate(year = 2020)

colnames(happiness_2017) = c("Country", "Happiness.Rank", "Happiness.Score",
                             "Upper.Confidence.Interval", "Lower.Confidence.Interval",
                             "Economy..GDP.per.Capita.", "Family", "Health..Life.Expectancy.",
                             "Freedom", "Generosity", "Trust..Government.Corruption.",
                             "Dystopia.Residual", "year")

colnames(happiness_2018) = c("Happiness.Rank", "Country", "Happiness.Score",
                             "Economy..GDP.per.Capita.", "Family", "Health..Life.Expectancy.",
                             "Freedom", "Generosity", "Trust..Government.Corruption.",
                             "year")

colnames(happiness_2019) = c("Happiness.Rank", "Country", "Happiness.Score",
                             "Economy..GDP.per.Capita.", "Family", "Health..Life.Expectancy.",
                             "Freedom", "Generosity", "Trust..Government.Corruption.",
                             "year")

colnames(happiness_2020) = c("Country", "Region", "Happiness.Score", "Standard.Error",
                             "Upper.Confidence.Interval", "Lower.Confidence.Interval",
                             "Logged.GDP.per.capita", "Family", "Healthy.life.expectancy",
                             "Freedom", "Generosity", "Trust..Government.Corruption.",
                             "Ladder.score.in.Dystopia", "Explained.by..Log.GDP.per.capita",
                             "Explained.by..Social.support",
                             "Explained.by..Healthy.life.expectancy",
                             "Explained.by..Freedom.to.make.life.choices",
                             "Explained.by..Generosity",
                             "Explained.by..Perceptions.of.corruption",
                             "Dystopia.Residual", "year")

happiness_2018 <- happiness_2018[-20,]
happiness_2018 <- happiness_2018 %>% mutate(Trust..Government.Corruption. =
                                              as.numeric(Trust..Government.Corruption.))

happiness_2015 <- happiness_2015 %>% mutate(mean = mean(Happiness.Score))
happiness_2016 <- happiness_2016 %>% mutate(mean = mean(Happiness.Score))
happiness_2017 <- happiness_2017 %>% mutate(mean = mean(Happiness.Score))
happiness_2018 <- happiness_2018 %>% mutate(mean = mean(Happiness.Score))
happiness_2019 <- happiness_2019 %>% mutate(mean = mean(Happiness.Score))
happiness_2020 <- happiness_2020 %>% mutate(mean = mean(Happiness.Score))

happiness <- happiness_2015 %>% bind_rows(happiness_2016) %>% bind_rows(happiness_2017) %>%
  bind_rows(happiness_2018) %>% bind_rows(happiness_2019) %>% bind_rows(happiness_2020)

#joining together happiness and country_code dataset
happiness_plotly <- left_join(country_code, happiness, by = c("English short name lower case" = "Country"))

#renaming the happiness dataset variables
colnames(happiness_plotly) = c("Country", "two.code", "three.code", "Numeric.code", "ISO",
                        "Region", "Happiness.Rank", "Happiness.Score", "Standard.Error",
                        "Economy..GDP.per.Capita.", "Family", "Health..Life.Expectancy.",
                        "Freedom", "Trust..Government.Corruption.", "Generosity",
                        "Dystopia.Residual", "year", "mean", "Lower.Confidence.Interval",
                        "Upper.Confidence.Interval", "Logged.GDP.per.capita",
                        "Healthy.life.expectancy", "Ladder.score.in.Dystopia",
                        "Explained.by..Log.GDP.per.capita", "Explained.by..Social.support",
                        "Explained.by..Healthy.life.expectancy",
                        "Explained.by..Freedom.to.make.life.choices",
                        "Explained.by..Generosity",
                        "Explained.by..Perceptions.of.corruption")

#"hover" variable: this specifies what info is displayed when you hover over a country
happiness_plotly <- happiness_plotly %>% mutate(hover = paste0(Country, "\nHappiness Rank: ", Happiness.Rank, "\nHappiness Score: ", Happiness.Score))

#hover label aesthetics
fontStyle <- list(family = "Arial", size = 15, color = "black")
label <- list(bgcolor = "#EEEEEE", bordercolor = "transparent", font = fontStyle)

#plotly map!
p1 <- plot_geo(happiness_plotly, locationmode = "world", frame = ~year) %>%
  add_trace(locations = ~three.code,
            z = ~Happiness.Score,
            zmin = 2,
            zmax = 8,
            color = ~Happiness.Score,
            colorscale = "Electric", #changes color of scale (other options: RdBu, Viridis, Rainbow)
            text = ~hover, #displays info in "hover" variable we made above
            hoverinfo = "text") %>% #makes hover info the text from above line
  layout(font = list(family = "Arial"), title = "World Happiness Report \n 2015 - 2020") %>%
  style(hoverlabel = label) %>%
  config(displayModeBar = FALSE) #hides the plotly toolbar in top right

p1

# credit to this tutorial: https://www.youtube.com/watch?v=RrtqBYLf404
```

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 5, dpi = 200}
ggplot(happiness, aes(y = Region, x = Happiness.Score, fill = Region)) + 
  geom_density_ridges(alpha = .8) +
  scale_fill_viridis_d(end = .75, option = "C") + 
  scale_color_viridis_d(end = .75, option = "C") + 
  theme_proj +
  theme(legend.position = "none") +
  labs(title = "Happiness Score of Each Region",
       x = "Happiness Score",
       y = "Region",
       caption = "SOURCE: World Happiness Report")
```

---
# Exploring Crime Dataset

(Blank for now!)

---
# Crime & Happiness Over Time

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 5, dpi = 200}
ggplot(data = happiness) +
  geom_point(mapping = aes(x = year, y = mean)) +
  geom_line(mapping = aes(x = year, y = mean)) +
  labs(title = "World Happiness Score Over Time",
       subtitle = "average of every country's score each year",
       x = "",
       y = "Happiness Score",
       caption = "SOURCE: World Happiness Report") +
  theme_proj +
  #need to add gridlines because tufte removes them
  theme(panel.grid.major = element_line(color = "gray90"),
        panel.grid.minor = element_line(color = "gray90"))
```

(Insert crime plots here?)

---
# Comparisons

```{r, echo = FALSE, message = FALSE, warning = FALSE, fig.height = 4, fig.width = 5, dpi = 200}
spotify_by_year_small <- spotify_by_year %>% filter(year == 2015 | year == 2016 | year == 2017 | year == 2018 | year == 2019 | year == 2020)

spotify_and_happiness <- left_join(spotify_by_year_small, happiness, by = c("year" = "year"))
ggplot(data = spotify_and_happiness) +
  geom_point(mapping = aes(x = valence, y = mean)) +
  theme_proj +
  #need to add gridlines because tufte removes them
  theme(panel.grid.major = element_line(color = "gray90"),
        panel.grid.minor = element_line(color = "gray90"))
```

(Insert crime & music comparison plot here)

---
# Discussion

